Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-14 introduction.tex
revisit exemples (ex. sw-VPA pas termine)

The latter case is related to a motivation  of  the present work:
automated music transcription. Representations that capture  music performances
are essentially linear:   
audio files, or the widely used MIDI file format~\cite{MIDIfile}.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in Common Western Music Notation~\cite{Gould11Notation}
%in music notation~\cite{Gould11Notation}: 
music scores are partitioned in measures, 
measures in beats, and beats can be further recursively divided.
It follows that music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the  timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear representation (audio or MIDI) and aims at re-constructing
these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}
over an infinite alphabet of timed events.

@@

\begin{example}[Running example]\label{ex:running}
We illustrate our framework with music transcription: given
a \emph{timeline} of timestamped musical events from an infinite
alphabet $\Sigma$ as input, 
parse it into a structured music score. 
Input events of $\Sigma$
are pairs $\< \mu, \tau>$, were $\mu$ is a MIDI message symbol~\cite{MIDIfile}
and $\tau \in \mathbb{Q}$ is a timestamp. Such inputs typically correspond
to the recording of a live performance. The output of parsing
is a sequence of events (or \emph{notes}) $\< \nu, \tau'>$ 
in an alphabet $\Delta$
%in Common Western Music Notation (CWMN)~\cite{Gould11Notation}
where event symbols $\nu$ are \emph{pitch names} (e.g., A4, G5, \etc.), 
and timestamp $\tau'$ belongs to a rhythmic ``grid'' obtained from recursive divisions:
whole notes ($\musWhole$) splitted in halves ($\musHalf$), halves
in quarters ($\musQuarter$), eights ($\musEighth$), etc.
The following concrete sequences
will be used:
\begin{enumerate}
  \item A performance  $I = [ \<49, 0.07>, \<51, 0.72>, \<52, 0.91>, \<54, 1.05>, \<56, 1.36>, \<57, 1.71>]$, % over interval $[0,2[$,
  \item A score $O =$ 
  \includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex1.pdf},
  %\includegraphics[scale=0.20]{pictures/score5.png}, 
  corresponding to a hierarchical structures (measures, tuplets) 
  that can be linearized as a sequence of events
      $[\<\mathrm{A}4, 0>, 
        \<\mathrm{B}4, \frac{3}{4}>, 
        \<\mathrm{C}5, \frac{7}{8}>, 
        \<\mathrm{D}5, 1>, 
        \<\mathrm{E}5,\frac{4}{3}>, 
        \<\mathrm{F}5\frac{5}{3}>]$
\end{enumerate}
\smallskip

We will show that $O$ is a soution for the
parsing of $I$. Note that several other parsings (e.g.,
\includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex2.pdf}) are possible.
SW-parsing associates a weight value
to each solution, and our framework
aims at selecting the best one with respect to this weight.
\endex
\end{example}

>>>>

The latter case is related to a motivation  of  the present work:
automated music transcription. Representations that capture  music performances
are essentially linear:   
audio files, or the widely used 
MIDI format~\cite{Selfridge-Field97beyondMIDI}. %\cite{MIDIfile}.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in Common Western Music Notation~\cite{Gould11Notation}
%in music notation~\cite{Gould11Notation}: 
music scores are partitioned in measures, 
measures in beats, and beats can be further recursively divided.
It follows that music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the  timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear representation (audio or MIDI) and aims at re-constructing
these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}
over an infinite alphabet of timed events.

@@

\begin{example}%[Running example]
\label{ex:running}
We illustrate our framework with a very simplified %toy
running example of \emph{music transcription}: 
a given \emph{timeline} of timestamped musical events 
from an infinite alphabet $\Sigma$ as input, 
is parsed into a structured music score. 
Input events of $\Sigma$
are pairs $\mu \ts{\tau}$, where $\mu$ is a 
MIDI pitch~\cite{Selfridge-Field97beyondMIDI}, %\cite{MIDIfile}
and $\tau \in \mathbb{Q}$ is a timestamp in seconds. 
Such inputs typically correspond
to the recording of a live performance, \eg
$I = 69\ts{0.07}, 
     71\ts{0.72}, 
     72\ts{0.91}, 
     74\ts{1.05}, 
     76\ts{1.36}, 
     77\ts{1.71}$. % over interval $[0,2[$,

The output of parsing is a sequence of 
timed symbols $\nu\ts{\tau'}$ in an alphabet $\Delta$,
%events (or \emph{notes}) 
%in Common Western Music Notation (CWMN)~\cite{Gould11Notation}
where every symbol $\nu$ may represent 
an \emph{event} (or \emph{note}), 
specified by its \emph{pitch name} 
(\eg, $\mbox{\footnotesize A4}$, $\mbox{\footnotesize G5}$, \etc.), 
an event \emph{continuation} (symbol `$-$`, see Ex.~\ref{ex:SWT}),
or a \emph{markup}. %(\emph{parenthesis}),
The timestamp $\tau'$, expressed in number of \emph{measures},
belongs to a rhythmic ``grid'' obtained from recursive divisions:
whole notes ($\musWhole$) split in halves ($\musHalf$), halves
in quarters ($\musQuarter$), eights ($\musEighth$), \etc.
%
For instance, the output score 
\includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex1.pdf},
  %\includegraphics[scale=0.20]{pictures/score5.png}, 
corresponds to a hierarchical structure 
(measures, tuplets..., see Figure~\ref{fig:score-tree}) 
that can be linearized as the sequence
$O =$ 
$\ccall{\mathsf{m}}\ts{0}$,
$\ccall{2}\ts{0}$,
$\mbox{\footnotesize A4}\ts{0}$, 
$\ccall{2}\ts{\frac{1}{2}}$,
$-\ts{\frac{1}{2}}$,
$\ccall{2}\ts{\frac{3}{4}}$,
$\mbox{\footnotesize B4}\ts{\frac{3}{4}}$, 
$\mbox{\footnotesize C$\sharp$5}\ts{\frac{7}{8}}$, 
$\creturn{2}\ts{1}$,
$\creturn{2}\ts{1}$,
$\creturn{2}\ts{1}$,
$\ccall{\mathsf{m}}\ts{1}$,
$\ccall{3}\ts{1}$,
$\mbox{\footnotesize D5}\ts{1}$, 
$\mbox{\footnotesize E5}\ts{\frac{4}{3}}$, 
$\mbox{\footnotesize F5}\ts{\frac{5}{3}}$,
$\creturn{3}\ts{2}$,
$\creturn{\mathsf{m}}\ts{2}$,
$\creturn{\mathsf{m}}\ts{2}$.
%\end{enumerate}
The opening markups $\ccall{\mathsf{m}}$ %and $\creturn{\mathsf{m}}$
delimit \emph{measures}, 
which are time intervals of duration 1 in this example, 
while the subsequences of $O$ between markups 
$\ccall{d}$ and $\creturn{d}$,
for some natural number $d$, 
represent a division of the current 
time interval into $d$ sub-intervals of equal length.

\smallskip
We will show that $O$ is a solution for the
parsing of $I$. Note that several other parsings (e.g.,
\includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex2.pdf}) are possible.
SW-parsing associates a weight value
to each solution, and our framework
aims at selecting the best one with respect to this weight.
\endex
\end{example}

### 2021-07-14 main.tex
ex.1 introduction
revisit exemples (ex. sw-VPA pas termine)

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SW{\textsf{sw}\xspace}
\def\swM{\textsf{swM}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\VPA{\textsf{VPA}\xspace}
\def\SVPA{\textsf{sVPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\newcommand{\weie[1]}{\wei_{#1}^\mathsf{e}}
\def\weiei{\weie[\mathsf{i}]}
\def\weiec{\weie[\mathsf{c}]}
\def\weier{\weie[\mathsf{r}]}
\def\weiex{\weie[\mathsf{x}]}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}

\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\newcommand{\ccall}[1]{\ensuremath \langle_{#1}}
\newcommand{\creturn}[1]{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$

@@

\begin{example}\label{ex:SWT}
We build a small weighted transducer model
with two states $q_0$ and $q_1$ that
calculates the distance between the performance $I$ and
the score $O$ of Example~\ref{ex:running}.

If one performed event $\mu_i$  corresponds
to one notated event $\nu_j$ (for instance MIDI code 61 and pitch A4),
the weight value computed by the \SWT is the time distance between both,
(see Example~\ref{distance-time}), and is modeled by
transitions $\wei_{11}$ below.
%
If we meet the music notation symbol '-' that
represents continuation (this occurs for instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{}), it is  skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%transitions
\[
\begin{array}{rclcrcl}
\wei_{11}(q_0, \<\mu, \tau>, \<\nu, \tau'>, q_0) & = & |\tau' - \tau| & \quad &
\wei_{11}(q_1, \<\mu, \tau>, \<\nu, \tau'>, q_0) & = & |\tau' - \tau|\\
\wei_{01}(q_0, \varepsilon, \< \mathsf{-}, \tau'>, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \< \mathsf{-}, \tau'>, q_0) & = & \one\\
\wei_{10}(q_0, \<\mu, \tau>, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also want to take performing errors into account, while still being able to compare with the score,
since a performer could, for example, play an unwritten extra note.
\lydia{reformulated this sentence}
%
This is modelled by the transition $\wei_{10}$ with an arbitrary weight value $\alpha \in \Semiring$,
switching from state $q_0$ (normal) to $q_1$ (error).
\philippe{Comprends pas cette phrase}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.
This \SWT is capable of evaluating the differences between a score and a performance,
all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

\begin{example}\label{ex:nested-word}
Consider once more the scores of Example~\ref{ex:running}.
$\Deltai$ represents the set of music events of the form $<\nu, \tau>$
where $\nu$ is either a note (e.g. A4), the continuation symbol $-$, or a
rest symbol R. $\Deltac$ and  $\Deltar$ are used to encode the score structure:
Let $\Deltac = \{\langle, \prec\}$ and  $\Deltar = \{\rangle, \succ\}$ where $\langle$
(resp. $\rangle$)
and $\prec$ (resp. $\succ$) correspond to start/end of measures, and start/end of
tuplets. Then  \includegraphics[scale=0.20]{pictures/score5.png}
is encoded with the nested word $\langle \prec 0, \frac{3}{4}, \frac{7}{8} \succ \rangle \langle \prec 1, \frac{4}{3}, \frac{5}{3}  \succ \rangle$
\endex
\end{example}

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA)
over  $\Delta = \Deltai \uplus \Deltac \uplus \Deltar$, $\Semiring$ and $\bar\Phi$
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states,
$P$ is a finite set of stack symbols,
$\mathsf{in} : Q \to \Semiring$
(respectively $\mathsf{out} : Q \to \Semiring$)
are functions defining the weight for entering
(respectively leaving) a state,
and $\bar\wei$ is a sextuplet composed of the transition functions :
$\weii : Q \times P \times Q \to \Phici$,
$\weiei : Q \times Q \to \Phii$,
$\weic : Q \times P \times Q \times P \to \Phicc$,
$\weiec : Q \times P \times Q \to \Phic$,
$\weir : Q \times P \times Q \to \Phicr$,
$\weier : Q \times Q \to \Phir$.
%and
%$\weiex : Q \times Q \to \Phix$
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
Similarly as in Section~\ref{section:transducer},
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$,
$a \in \Deltai$,
$\call{c} \in \Deltac$,
$\return{r} \in \Deltar$,
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times [\Deltac \times P] \times \Deltai \times Q \to \Semiring &
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) &
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Deltai \times Q \to \Semiring &
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\
%
\weic: Q \times [\Deltac \times P] \times  [\Deltac \times P] \times Q \to \Semiring &
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) &
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times [\Deltac \times P] \times Q \to \Semiring &
\weiec(q, {\call{c}}, p, q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\
%
\weir: Q \times [\Deltac \times P] \times \Deltar \times Q \to \Semiring &
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) &
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Deltar \times Q \to \Semiring &
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}
\]

@@

@@

\begin{example}[Symbolic Weighted Parsing and the transcription problem]
Applied to the music transcription problem, the above formalism
is interpreted as follows:
\begin{enumerate}

  \item The \SWT $T$ evaluates a ``fitness measure''  that expresses
    a correspondance between a performance and its notation. See Example~\ref{ex:SWT}.
\item The \SWVPA $A$ expresses a cost related to the music notation. One possibility
 is to relate this cost to the structural complexity: given a set of equivalent
 representations, we aim at choosing the simpler one. For instance
 \includegraphics[scale=0.20]{pictures/score5.png} should be favored
 on \includegraphics[scale=0.20]{pictures/score4.png}. Costs  can be expressed
 in the grammar as weights associated to each production rule.
 For instance, the following production rules define two possible divisions
 of a time interval into respectively a duplet and a triplet. Each comes with
 a specific cost.
 \[
 \rho_1: q_0 \xrightarrow{0.06} \< q_{1}, q_{2}>,\
 \rho_2: q_0 \xrightarrow{0.12} \< q_{1}, q_{2}, q_{2}>.
 \]
 Further binary divisions of time sub-intervals %represented respectively by~$q_2$ and~$q_3$
 are possible with:
 \[
 \rho_3:\,q_2 \xrightarrow{0.1} \< q_{3}, q_{3}>, \
 \rho_4:\,q_3 \xrightarrow{0.11} \< q_{4}, q_{4}>. \
 \]

Score \includegraphics[scale=0.20]{pictures/score5.png} is obtained as follows:
the first measure
 results from the successive applications of  rules $\rho_1$ (division in two)
 and rules $\rho_3$ (division of the second half) in two. The second measure
 is a division in three obtained by rule $\rho_3$. One obtains a parse tree
 that can be linearized in the nested word of Example~\ref{ex:nested-word}.
 \end{enumerate}
Therefore, the framework, applied to the transcription problem, allows to find an optimal
solution that considers both the fitness of the result, and its structural complexity.
\end{example}

>>>>

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SW{\textsf{sw}\xspace}
\def\swM{\textsf{swM}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\VPA{\textsf{VPA}\xspace}
\def\SVPA{\textsf{sVPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\newcommand{\weie[1]}{\wei_{#1}^\mathsf{e}}
\def\weiei{\weie[\mathsf{i}]}
\def\weiec{\weie[\mathsf{c}]}
\def\weier{\weie[\mathsf{r}]}
\def\weiex{\weie[\mathsf{x}]}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}
\newcommand{\ts}[1]{\mathop{:}{#1}} % timestamp marker
%\newcommand{\ts}[1]{^{#1}} % timestamp marker

\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\newcommand{\ccall}[1]{\ensuremath \langle_{#1}}
\newcommand{\creturn}[1]{\ensuremath \rangle_{#1}} 
%\newcommand{\creturn}[1]{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$

@@

\begin{example}\label{ex:SWT}
We build a small \SWT over the alphabets $\Sigma$ and $\Deltai$ of 
Examples~\ref{ex:running} and \ref{distance-time},
%weighted transducer model
with two states $q_0$ and $q_1$ that
calculates the temporal distance between an input performance in~$\Sigma^*$
and the subsequence 
of the events of $\Deltai$ in a score.
%the performance $I$ and
%the score $O$ of Example~\ref{ex:running}.
%
When reading one performed event $\mu$ 
and the corresponding notated event $\nu$ 
(for instance MIDI pitch 69 and note name $\mbox{\footnotesize A4}$),
the weight value computed by the \SWT is the time distance between both,
(see Example~\ref{distance-time}), as modeled by transitions $\wei_{11}$ below.
%
If we meet the continuation symbol $'-'$ 
(this occurs for instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{}),
%that represents continuation 
it is  skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%Likewise, the markups are ignored.
\[
\begin{array}{lclclcll}
\wei_{11}(q_0, \mu\ts{\tau}, \nu\ts{\tau'}, q_0) & = & 
     \delta(\mu\ts{\tau}, \nu\ts{\tau'}) & \quad & %|\tau' - \tau| & \quad &
\wei_{11}(q_1, \mu\ts{\tau}, \nu\ts{\tau'}, q_0) & = & 
     \delta(\mu\ts{\tau}, \nu\ts{\tau'}) & 
     \mbox{if~} \nu \neq - \\ %|\tau' - \tau| & \quad &
\wei_{01}(q_0, \varepsilon, \mathsf{-}\ts{\tau'}, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \mathsf{-}\ts{\tau'}, q_0) & = & \one\\
\wei_{10}(q_0, \mu\ts{\tau}, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also want to take performing errors into account, 
while still being able to compare with the score,
since a performer could, for example, play an unwritten extra note.
%\lydia{reformulated this sentence}
%
%Reading such a note is modelled by 
The transition $\wei_{10}$, 
with an fixed weight value $\alpha \in \Semiring$,
switches from state $q_0$ (normal) to $q_1$ (error)
when reading an extra note $\mu$.
\philippe{Comprends pas cette phrase}
\florent{$\to$ reformulation}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.
\florent{dernière phrase utile?}
This \SWT is capable of evaluating the differences between a score and a performance,
all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

\begin{example}\label{ex:nested-word}
Consider once more the linearized score of Example~\ref{ex:running},
with $\Deltai$ representing timed notes and continuations.
%the set of music events of the form $<\nu, \tau>$
%where $\nu$ is either a note (e.g. A4), the continuation symbol $-$, or a
%rest symbol R. 
The subset $\Delta \setminus \Deltai$ is partitioned into 
$\Deltac$ and $\Deltar$, containing respectively opening and closing parentheses.
%are used to encode the score structure:
%Let $\Deltac = \{\langle, \prec\}$ and  $\Deltar = \{\rangle, \succ\}$ where $\langle$
%(resp. $\rangle$)
%and $\prec$ (resp. $\succ$) correspond to start/end of measures, and start/end of
%tuplets. Then  \includegraphics[scale=0.20]{pictures/score5.png}
%is encoded with the nested word $\langle \prec 0, \frac{3}{4}, \frac{7}{8} \succ \rangle %\langle \prec 1, \frac{4}{3}, \frac{5}{3}  \succ \rangle$
The score representation $O$ in Example~\ref{ex:running}
% score: \includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex1.pdf},
is a nested-word of~$\Delta^*$. 
%
Another example is the second candidate 
\includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex2.pdf}
of transcription of~$I$,
linearized into
$O' =$ 
$\ccall{\mathsf{m}}\ts{0}$,
$\ccall{2}\ts{0}$,
$\mbox{\footnotesize A4}\ts{0}$, 
$\ccall{2}\ts{\frac{1}{2}}$,
$-\ts{\frac{1}{2}}$,
$\mbox{\footnotesize B4}\ts{\frac{3}{4}}$, 
$\creturn{2}\ts{1}$,
$\creturn{2}\ts{1}$,
$\ccall{\mathsf{m}}\ts{1}$,
$\ccall{3}\ts{1}$,
$\mbox{\footnotesize `C$\sharp$5`}\ts{1}$, 
$\mbox{\footnotesize D5}\ts{1}$, 
$\mbox{\footnotesize E5}\ts{\frac{4}{3}}$, 
$\mbox{\footnotesize F5}\ts{\frac{5}{3}}$,
$\creturn{3}\ts{2}$,
$\creturn{\mathsf{m}}\ts{2}$,
$\creturn{\mathsf{m}}\ts{2}$
(see also Figure~\ref{fig:score-tree}).
The symbol between quotes $\mbox{\footnotesize `C$\sharp$5`}$
represents an \emph{appogiatura}, %(grace-note),
that is an ornemental note with theoretical duration zero.
\endex
\end{example}

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA)
over  $\Delta = \Deltai \uplus \Deltac \uplus \Deltar$, $\Semiring$ and $\bar\Phi$
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states,
$P$ is a finite set of stack symbols,
$\mathsf{in} : Q \to \Semiring$
(respectively $\mathsf{out} : Q \to \Semiring$)
are functions defining the weight for entering
(respectively leaving) a state,
and $\bar\wei$ is a sextuplet composed of the transition functions :
$\weii : Q \times P \times Q \to \Phici$,
$\weiei : Q \times Q \to \Phii$,
$\weic : Q \times P \times Q \times P \to \Phicc$,
$\weiec : Q \times P \times Q \to \Phic$,
$\weir : Q \times P \times Q \to \Phicr$,
$\weier : Q \times Q \to \Phir$.
%and
%$\weiex : Q \times Q \to \Phix$
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
Similarly as in Section~\ref{section:transducer},
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$,
$a \in \Deltai$,
$\call{c} \in \Deltac$,
$\return{r} \in \Deltar$,
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times [\Deltac \times P] \times \Deltai \times Q \to \Semiring &
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) &
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Deltai \times Q \to \Semiring &
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\
%
\weic: Q \times [\Deltac \times P] \times  [\Deltac \times P] \times Q \to \Semiring &
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) &
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times [\Deltac \times P] \times Q \to \Semiring &
\weiec(q, {\call{c}}, p', q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\
%
\weir: Q \times [\Deltac \times P] \times \Deltar \times Q \to \Semiring &
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) &
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Deltar \times Q \to \Semiring &
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}
\]

@@

\begin{example}
... \SWVPA over the alphabet of Example~\ref{ex:nested-word}
that expresses a weight related to the music notation,
more precisely to its structural complexity: 
given a set of equivalent representations, we aim at choosing the simpler one. 
...

*** SUPPR ***
For instance
\includegraphics[scale=0.20]{pictures/score5.png} will be favored
on \includegraphics[scale=0.20]{pictures/score4.png}. 
%
Costs  can be expressed
in the grammar as weights associated to each production rule.
For instance, the following production rules define two possible divisions
of a time interval into respectively a duplet and a triplet. Each comes with
a specific cost.
\[
\rho_1: q_0 \xrightarrow{0.06} \< q_{1}, q_{2}>,\
\rho_2: q_0 \xrightarrow{0.12} \< q_{1}, q_{2}, q_{2}>.
\]
Further binary divisions of time sub-intervals %represented respectively by~$q_2$ and~$q_3$
are possible with:
\[
\rho_3:\,q_2 \xrightarrow{0.1} \< q_{3}, q_{3}>, \
\rho_4:\,q_3 \xrightarrow{0.11} \< q_{4}, q_{4}>. \
\]

Score \includegraphics[scale=0.20]{pictures/score5.png} is obtained as follows:
the first measure
results from the successive applications of  rules $\rho_1$ (division in two)
and rules $\rho_3$ (division of the second half) in two. The second measure
is a division in three obtained by rule $\rho_3$. One obtains a parse tree
that can be linearized in the nested word of Example~\ref{ex:nested-word}.
...
\end{example}

@@

\begin{example}[Symbolic Weighted Parsing and the transcription problem]
Applied to the music transcription problem, the above formalism
is interpreted as follows:
%
\begin{itemize}
\item The input word is $I$ of Example~\ref{ex:running}.
\item The \SWT $T$ evaluates a ``fitness measure''  that expresses
    a correspondance between a performance and notation events. See Example~\ref{ex:SWT}.
\item The \SWVPA $A$ expresses a cost related to the music notation. 
\end{itemize}
One possibility is to relate this cost to the structural complexity: 
given a set of equivalent representations, 
we aim at choosing the simpler one. 
% 
For instance
\includegraphics[scale=0.20]{pictures/score5.png} will be favored
on \includegraphics[scale=0.20]{pictures/score4.png}. Costs  can be expressed
in the grammar as weights associated to each production rule.
For instance, the following production rules define two possible divisions
of a time interval into respectively a duplet and a triplet. Each comes with
a specific cost.
\[
\rho_1: q_0 \xrightarrow{0.06} \< q_{1}, q_{2}>,\
\rho_2: q_0 \xrightarrow{0.12} \< q_{1}, q_{2}, q_{2}>.
\]
Further binary divisions of time sub-intervals %represented respectively by~$q_2$ and~$q_3$
are possible with:
\[
\rho_3:\,q_2 \xrightarrow{0.1} \< q_{3}, q_{3}>, \
\rho_4:\,q_3 \xrightarrow{0.11} \< q_{4}, q_{4}>. \
\]

Score \includegraphics[scale=0.20]{pictures/score5.png} is obtained as follows:
the first measure
results from the successive applications of  rules $\rho_1$ (division in two)
and rules $\rho_3$ (division of the second half) in two. The second measure
is a division in three obtained by rule $\rho_3$. One obtains a parse tree
that can be linearized in the nested word of Example~\ref{ex:nested-word}.

Therefore, the framework, applied to the transcription problem, allows to find an optimal
solution that considers both the fitness of the result, and its structural complexity.
\end{example}
